## Week 5 Formative Assessment

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)

options(scipen = 999)

library(ggplot2)
library(psych)
library(knitr)
```



# Import the data

```{r import-data}
# Read the CSV from the root folder
job_data <- read.csv("data/job_survey_data.csv")

head(job_data)
```



# Check the data

```{r check-data}
# Check the dataset
cat("Rows:", nrow(job_data), "\n")
cat("Columns:", ncol(job_data), "\n")
cat("Missing values:", sum(is.na(job_data)), "\n")
cat("Duplicate rows:", sum(duplicated(job_data)), "\n")

names(job_data)
```



# Question 1: The absence variable

```{r absence-summary}
# Summarise absence
absence_summary <- data.frame(
  Statistic = c(
    "Valid cases",
    "Mean",
    "Median",
    "Standard deviation",
    "Minimum",
    "Maximum"
  ),
  Value = c(
    sum(!is.na(job_data$absence)),
    mean(job_data$absence, na.rm = TRUE),
    median(job_data$absence, na.rm = TRUE),
    sd(job_data$absence, na.rm = TRUE),
    min(job_data$absence, na.rm = TRUE),
    max(job_data$absence, na.rm = TRUE)
  )
)

kable(
  absence_summary,
  digits = 2,
  caption = "Descriptive statistics for absence"
)
```

```{r absence-frequency}
# Create the frequency table
absence_frequency <- as.data.frame(
  table(job_data$absence, useNA = "ifany")
)

names(absence_frequency) <- c("Absence", "Frequency")

kable(
  absence_frequency,
  caption = "Frequency distribution of absence"
)
```

```{r absence-frequency-plot}
# Plot the absence frequencies
ggplot(absence_frequency, aes(x = factor(Absence), y = Frequency)) +
  geom_col(fill = "steelblue", colour = "black") +
  labs(
    title = "Frequency of Employee Absence",
    x = "Absence",
    y = "Number of employees"
  ) +
  theme_minimal()
```

```{r absence-boxplot}
# Check the value 99
ggplot(job_data, aes(y = absence)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Box Plot of Employee Absence",
    x = NULL,
    y = "Absence"
  ) +
  theme_minimal()
```

```{r absence-sensitivity}
# Compare results with and without 99
absence_without_99 <- job_data$absence[job_data$absence != 99]

absence_sensitivity <- data.frame(
  Analysis = c("All values", "Value 99 excluded"),
  N = c(
    length(job_data$absence),
    length(absence_without_99)
  ),
  Mean = c(
    mean(job_data$absence),
    mean(absence_without_99)
  ),
  Median = c(
    median(job_data$absence),
    median(absence_without_99)
  ),
  SD = c(
    sd(job_data$absence),
    sd(absence_without_99)
  ),
  Maximum = c(
    max(job_data$absence),
    max(absence_without_99)
  )
)

kable(
  absence_sensitivity,
  digits = 2,
  caption = "Sensitivity analysis for absence"
)
```



The value 99 should only be recoded after checking the codebook.

# Question 2: Test-retest reliability

Test-retest reliability is external reliability because the same measure is compared across separate occasions. The Job Survey does not contain repeated administrations, so no test-retest coefficient is calculated.

# Question 3: Cronbach's alpha for autonomy

```{r select-autonomy}
# Select the autonomy items
autonom <- job_data[
  c("autonom1", "autonom2", "autonom3", "autonom4")
]

str(autonom)
summary(autonom)
colSums(is.na(autonom))
```



The item wording is not available in the CSV. `check.keys = TRUE` is used as a diagnostic only. Reverse scoring should be confirmed from the questionnaire.

```{r autonomy-distributions}
# Reshape the autonomy items
long_autonom <- stack(autonom)
names(long_autonom) <- c("Score", "Item")

# Plot the item distributions
ggplot(long_autonom, aes(x = factor(Score))) +
  geom_bar(fill = "steelblue", colour = "black") +
  facet_wrap(~ Item, ncol = 2) +
  labs(
    title = "Distributions of Job-Autonomy Items",
    x = "Response score",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r autonomy-correlations}
# Calculate item correlations
autonomy_correlations <- cor(
  autonom,
  use = "pairwise.complete.obs"
)

kable(
  autonomy_correlations,
  digits = 2,
  caption = "Correlations between autonomy items"
)
```

```{r cronbach-alpha}
# Calculate Cronbach's alpha
alpha_result <- psych::alpha(
  autonom,
  check.keys = TRUE,
  warnings = FALSE
)

alpha_summary <- data.frame(
  Statistic = c(
    "Raw Cronbach's alpha",
    "Standardised Cronbach's alpha",
    "Number of items",
    "Complete cases"
  ),
  Value = c(
    alpha_result$total$raw_alpha,
    alpha_result$total$std.alpha,
    ncol(autonom),
    sum(complete.cases(autonom))
  )
)

kable(
  alpha_summary,
  digits = 3,
  caption = "Cronbach's alpha for the autonomy scale"
)
```

```{r item-total-correlations}
# Extract corrected item-total correlations
item_total <- data.frame(
  Item = rownames(alpha_result$item.stats),
  Corrected_Item_Total = alpha_result$item.stats$r.drop,
  row.names = NULL
)

kable(
  item_total,
  digits = 3,
  caption = "Corrected item-total correlations"
)
```

```{r item-total-plot}
# Plot corrected item-total correlations
ggplot(item_total, aes(x = Item, y = Corrected_Item_Total)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0.30, linetype = "dashed") +
  labs(
    title = "Corrected Item-Total Correlations",
    x = "Autonomy item",
    y = "Corrected item-total correlation"
  ) +
  theme_minimal()
```

```{r alpha-if-deleted}
# Extract alpha if each item is deleted
alpha_deleted <- data.frame(
  Item = rownames(alpha_result$alpha.drop),
  Alpha_If_Deleted = alpha_result$alpha.drop$raw_alpha,
  row.names = NULL
)

kable(
  alpha_deleted,
  digits = 3,
  caption = "Cronbach's alpha if each item is deleted"
)
```

```{r alpha-deleted-plot}
# Plot alpha if deleted
overall_alpha <- alpha_result$total$raw_alpha

ggplot(alpha_deleted, aes(x = Item, y = Alpha_If_Deleted)) +
  geom_col(fill = "steelblue") +
  geom_hline(
    yintercept = overall_alpha,
    colour = "red",
    linetype = "dashed"
  ) +
  labs(
    title = "Cronbach's Alpha if an Item Is Deleted",
    subtitle = "The dashed line shows the overall alpha",
    x = "Item removed",
    y = "Cronbach's alpha"
  ) +
  theme_minimal()
```



# Question 4: Alcohol-consumption variables

```{r measurement-levels}
# Summarise the measurement levels
measurement_levels <- data.frame(
  Variable = c(
    "Ever consumes alcohol: Yes or No",
    "Drink consumed most frequently",
    "Frequency of alcohol consumption",
    "Units consumed in the previous week"
  ),
  Level = c(
    "Dichotomous nominal",
    "Nominal",
    "Ordinal",
    "Ratio"
  ),
  Reason = c(
    "Two unordered categories",
    "Unordered categories",
    "Ordered categories with unequal intervals",
    "Equal units and a meaningful zero"
  )
)

kable(
  measurement_levels,
  caption = "Measurement levels of the alcohol-consumption variables"
)
```