## Week 5 Formative Assessment: Job Survey Analysis
## Gregory Antony Mikuro

The main aims are to:

1. identify the level of measurement of `absence`;
2. explain test-retest reliability;
3. assess the internal consistency of the job-autonomy items; and
4. classify the variables in the questions about alcohol consumption.

# Install required packages

Run this chunk once if the packages are not already installed. After installation, change `eval=FALSE` back to `eval=FALSE` before knitting the final report.

```{r install-packages, eval=FALSE}
install.packages("ggplot2")
install.packages("psych")
install.packages("knitr")
```




# Load required packages

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(psych)
library(knitr)

# Make tables easier to read
options(scipen = 999)
```




# Import the data

The CSV file is stored in the `data` folder inside the project folder.

```{r import-data}
# Read the Job Survey data
job_data <- read.csv("data/job_survey_data.csv")

# Show the first six rows
head(job_data)
```




# Check the data

Before carrying out the analysis, the dataset is checked for its size, variable names, missing values and duplicate rows.

```{r check-data}
# Number of rows and columns
cat("Number of rows:", nrow(job_data), "\n")
cat("Number of columns:", ncol(job_data), "\n\n")

# Variable names
names(job_data)
```

```{r data-structure}
# Check the structure and data types
str(job_data)
```

```{r missing-and-duplicates}
# Count missing values in each variable
missing_values <- data.frame(
  Variable = names(job_data),
  Missing = colSums(is.na(job_data))
)

kable(missing_values, caption = "Missing values in the dataset")

# Count duplicate rows
cat("Number of duplicate rows:", sum(duplicated(job_data)))
```




# Question 1: The absence variable

## Descriptive statistics

```{r absence-summary}
# Calculate descriptive statistics for absence
absence_summary <- data.frame(
  Statistic = c("Valid cases", "Mean", "Median", "Standard deviation", "Minimum", "Maximum"),
  Value = c(
    sum(!is.na(job_data$absence)),
    mean(job_data$absence, na.rm = TRUE),
    median(job_data$absence, na.rm = TRUE),
    sd(job_data$absence, na.rm = TRUE),
    min(job_data$absence, na.rm = TRUE),
    max(job_data$absence, na.rm = TRUE)
  )
)

kable(absence_summary, digits = 2, caption = "Descriptive statistics for absence")
```




## Frequency table

```{r absence-frequency}
# Create a frequency table for absence
absence_frequency <- as.data.frame(table(job_data$absence, useNA = "ifany"))
names(absence_frequency) <- c("Absence", "Frequency")

kable(absence_frequency, caption = "Frequency distribution of absence")
```




## Histogram

```{r absence-histogram, fig.width=7, fig.height=4}
# Show the distribution of absence
ggplot(job_data, aes(x = absence)) +
  geom_histogram(bins = 10, fill = "steelblue", colour = "black") +
  labs(
    title = "Distribution of Employee Absence",
    x = "Absence",
    y = "Frequency"
  ) +
  theme_minimal()
```




## Box plot

```{r absence-boxplot, fig.width=6, fig.height=4}
# Check the spread and possible unusual values
ggplot(job_data, aes(y = absence)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Box Plot of Employee Absence",
    x = NULL,
    y = "Absence"
  ) +
  theme_minimal()
```

```{r select-autonomy}
# Select the four job-autonomy items
autonomy_data <- job_data[c("autonom1", "autonom2", "autonom3", "autonom4")]

head(autonomy_data)
```




## Check the autonomy values

```{r check-autonomy-values}
# Display the minimum and maximum for each item
range_check <- data.frame(
  Item = names(autonomy_data),
  Minimum = sapply(autonomy_data, min, na.rm = TRUE),
  Maximum = sapply(autonomy_data, max, na.rm = TRUE),
  Missing = colSums(is.na(autonomy_data))
)

kable(range_check, caption = "Range and missing-value check for autonomy items")
```





## Descriptive statistics

```{r autonomy-descriptives}
# Calculate simple descriptive statistics for each item
autonomy_summary <- data.frame(
  Item = names(autonomy_data),
  Mean = sapply(autonomy_data, mean, na.rm = TRUE),
  SD = sapply(autonomy_data, sd, na.rm = TRUE),
  Median = sapply(autonomy_data, median, na.rm = TRUE),
  Minimum = sapply(autonomy_data, min, na.rm = TRUE),
  Maximum = sapply(autonomy_data, max, na.rm = TRUE)
)

kable(autonomy_summary, digits = 2, caption = "Descriptive statistics for autonomy items")
```




## Autonomy item plots

```{r autonomy-plots, fig.width=8, fig.height=6}
# Reshape the data for plotting using base R
long_autonomy <- stack(autonomy_data)
names(long_autonomy) <- c("Score", "Item")

# Plot the frequency distribution of each item
ggplot(long_autonomy, aes(x = Score)) +
  geom_histogram(binwidth = 1, boundary = 0.5, fill = "steelblue", colour = "black") +
  facet_wrap(~ Item, ncol = 2) +
  labs(
    title = "Distributions of Job-Autonomy Items",
    x = "Response score",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r autonomy-boxplots, fig.width=7, fig.height=4}
# Compare the distributions of the autonomy items
ggplot(long_autonomy, aes(x = Item, y = Score)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Box Plots of Job-Autonomy Items",
    x = "Autonomy item",
    y = "Response score"
  ) +
  theme_minimal()
```




## Correlation matrix

```{r autonomy-correlations}
# Calculate correlations between the four items
# Pairwise complete observations are used if values are missing
autonomy_correlations <- cor(autonomy_data, use = "pairwise.complete.obs")

kable(autonomy_correlations, digits = 2, caption = "Correlations between autonomy items")
```

```{r correlation-plot, fig.width=6, fig.height=5}
# Draw a simple correlation plot
image(
  1:4,
  1:4,
  autonomy_correlations,
  axes = FALSE,
  col = colorRampPalette(c("red", "white", "blue"))(20),
  zlim = c(-1, 1),
  main = "Correlation Matrix of Autonomy Items"
)

axis(1, at = 1:4, labels = names(autonomy_data))
axis(2, at = 1:4, labels = names(autonomy_data), las = 1)
box()
```




## Cronbach's alpha

```{r cronbach-alpha}
# Remove rows with missing autonomy responses
autonomy_complete <- na.omit(autonomy_data)

# Calculate Cronbach's alpha
alpha_result <- psych::alpha(
  autonomy_complete,
  check.keys = TRUE,
  warnings = FALSE
)

# Display the main alpha results
alpha_table <- data.frame(
  Measure = c("Raw alpha", "Standardised alpha", "Number of items", "Complete cases"),
  Value = c(
    alpha_result$total$raw_alpha,
    alpha_result$total$std.alpha,
    ncol(autonomy_complete),
    nrow(autonomy_complete)
  )
)

kable(alpha_table, digits = 3, caption = "Cronbach's alpha for the autonomy scale")
```




## Corrected item-total correlations

```{r item-total-correlations}
# r.drop is the corrected item-total correlation
item_total <- data.frame(
  Item = rownames(alpha_result$item.stats),
  Corrected_Item_Total = alpha_result$item.stats$r.drop
)

kable(item_total, digits = 3, caption = "Corrected item-total correlations")
```

```{r item-total-plot, fig.width=7, fig.height=4}
# Plot the corrected item-total correlations
ggplot(item_total, aes(x = Item, y = Corrected_Item_Total)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0, colour = "black") +
  labs(
    title = "Corrected Item-Total Correlations",
    x = "Autonomy item",
    y = "Corrected item-total correlation"
  ) +
  theme_minimal()
```




## Alpha if an item is deleted

```{r alpha-if-deleted}
# Extract alpha after removing each item
alpha_deleted <- data.frame(
  Item = rownames(alpha_result$alpha.drop),
  Alpha_If_Deleted = alpha_result$alpha.drop$raw_alpha
)

kable(alpha_deleted, digits = 3, caption = "Cronbach's alpha if an item is deleted")
```

```{r alpha-deleted-plot, fig.width=7, fig.height=4}
# Compare alpha if an item is deleted with the overall alpha
overall_alpha <- alpha_result$total$raw_alpha

ggplot(alpha_deleted, aes(x = Item, y = Alpha_If_Deleted)) +
  geom_col(fill = "steelblue") +
  geom_hline(
    yintercept = overall_alpha,
    colour = "red",
    linetype = "dashed"
  ) +
  labs(
    title = "Cronbach's Alpha if an Item Is Deleted",
    subtitle = "The dashed red line is the overall alpha",
    x = "Item removed",
    y = "Cronbach's alpha"
  ) +
  theme_minimal()
```



# Question 4: Classification of alcohol-consumption variables

## Question 4(a)

"Do you ever consume alcoholic drinks?" is a dichotomous variable because it has only two possible categories: Yes and No.

## Question 4(b)

The type of alcoholic drink consumed most frequently is a nominal variable. Categories such as beer, spirits and wine do not have a natural order.

## Question 4(c)

Frequency of alcohol consumption is an ordinal variable. The response categories can be ordered from less frequent to more frequent consumption, but the gaps between the categories are not equal.

## Question 4(d)

The number of alcohol units consumed during the previous week is a ratio-level variable. It is numerical and has a meaningful zero, because zero means that no units were consumed.